---
title: "Housing Data"
output: html_notebook
---
The project aims to leverage data-driven insights to better understand housing market trends, predict property prices, and provide valuable information for potential home buyers and real estate investors.By harnessing the power of data analytics using R, we intend to see how the housing landscape
has changed over the last five years. Along the way, we also want to provide actionable
information by answering the below questions:
• Which Iowa counties have the highest and lowest property values?
• How have property values changed over time?
• What counties are similar based on some of the variables we used?
• How well does median income predict median housing value and vice versa?
• Does the racial makeup of a county affect the median housing value in that    area?

```{r}
# Load necessary libraries
library(readr)
library(tidyr)
library(ggplot2)
library(scales)
library(readxl)
library(cluster)
library(factoextra)
library(tidyverse)
library(ggmap)
library(choroplethr)
library(choroplethrMaps)
library(acs)
library(ggrepel)
library(tigris)

# Read the csv file for Income and Housing Unit
housing_df <- read_csv("Data/MedianCountyValue.csv", show_col_types = FALSE)

income_df <- read_csv("Data/IncomeHousing.csv", show_col_types = FALSE)

```
# Cleaning the data

```{r}

# Checking if there are missing values in the data frame

# Show records with missing values
records_with_missing_homeval <- housing_df[!complete.cases(housing_df), ]
records_with_missing_values <- income_df[!complete.cases(income_df), ]

# Filter out rows with missing values 
income_df <- income_df[complete.cases(income_df$`Measure Values`), ]

#Rename income dataframe county column name 
income_df <- income_df %>%
  rename(`Geographic name` = `Geographic Name`)


convert_dollar_to_int <- function(column) {
    
    indices <- grepl("^\\$", column)
    
   
    numeric_values <- rep(NA_real_, length(column))
    
 
    numeric_values[indices] <- as.numeric(gsub("[$,]", "", column[indices]))
    
  
    result <- ifelse(is.na(numeric_values), column, numeric_values)
    
    return(result)
}


for (col in names(housing_df)) {
    if (col != "Geographic name") {
        housing_df[[col]] <- convert_dollar_to_int(housing_df[[col]])
    }
}



```

# Q1) Which Iowa counties have the highest and lowest property values? 

```{r}
#Highest Property Values
highest_property_counties1 <- housing_df %>%
  filter(!is.na(`Estimate2017-2021`)) %>%
  arrange(desc(`Estimate2017-2021`)) %>%
  head(10)

# Lowest Property Values
lowest_property_counties1 <- housing_df %>%
  filter(!is.na(`Estimate2017-2021`)) %>%
  arrange(`Estimate2017-2021`) %>%
  head(10)  # Get the top 10 counties with lowest property values

#Highest Property Values
highest_property_counties2 <- housing_df %>%
  filter(!is.na(`Estimate2012-2016`)) %>%
  arrange(desc(`Estimate2012-2016`)) %>%
  head(10)

# Lowest Property Values
lowest_property_counties2 <- housing_df %>%
  filter(!is.na(`Estimate2012-2016`)) %>%
  arrange(`Estimate2012-2016`) %>%
  head(10)  # Get the top 10 counties with lowest property values

#Highest Property Values
highest_property_counties3 <- housing_df %>%
  filter(!is.na(`Estimate2007-2011`)) %>%
  arrange(desc(`Estimate2007-2011`)) %>%
  head(10)

# Lowest Property Values
lowest_property_counties3 <- housing_df %>%
  filter(!is.na(`Estimate2007-2011`)) %>%
  arrange(`Estimate2007-2011`) %>%
  head(10)  # Get the top 10 counties with lowest property values
```



```{r}

# Function to compute means for given columns
compute_means <- function(df, columns) {
  sapply(columns, function(col) mean(df[[col]]))
}

# Define the columns of interest
cols <- c("Estimate2007-2011", "Estimate2012-2016", "Estimate2017-2021")

# Compute means using the function
highmeans <- compute_means(highest_property_counties3, cols)
lowmeans <- compute_means(lowest_property_counties3, cols)

# Plot both lines
plot(highmeans, type = 'l', col = 'red', ylim = range(c(highmeans, lowmeans)), 
     xaxt = 'n', xlab = 'Time Period', ylab = 'Mean Value', main = 'Mean Value Over Time')
lines(lowmeans, col = 'blue')
axis(1, at = 1:3, labels = cols)
legend("topright", legend = c("High", "Low"), fill = c("red", "blue"), bty = "n")

```

```{r}
mygraph = plot(highmeans, type = "b", col = "blue", ylim = c(70000,250000), main = 'Change in Mean Housing Values from 2007-2021', ylab = 'Mean Value') 
lines(lowmeans, type = "b", col = "red") 

mygraph = mygraph + theme(axis.ticks.x = element_blank())
```

#Visualization
```{r}
# Highest Property Values: Bar Plot

p <- qplot(data = highest_property_counties1, x = reorder(`Geographic name`, -`Estimate2017-2021`), 
      y = `Estimate2017-2021`, geom = "col", fill = I("skyblue"),
      main = "Top 10 Counties with Highest Property Values",
      xlab = "County", ylab = "Property Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$"))

p

# Lowest Property Values: Bar Plot with y-axis labels showing numbers with $ sign
q <- qplot(data = lowest_property_counties1, x = reorder(`Geographic name`, `Estimate2017-2021`), 
            y = `Estimate2017-2021`, geom = "col", fill = I("salmon"),
            main = "Top 10 Counties with Lowest Property Values",
            xlab = "County", ylab = "Property Value") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(labels = scales::dollar_format(prefix = "$"))
q
```
# Q2) How have property values changed over time?

```{r}
# Function to Plot Property Value Trend for a Specific County using qplot
plot_property_value <- function(data, county_name) {
  county_data <- data %>%
    filter(`Geographic name` == county_name) %>%
    gather(key = "year_range", value = "property_value", 
           `Estimate2017-2021`, `Estimate2012-2016`, `Estimate2007-2011`)
  
  qplot(data = county_data, x = as.factor(year_range), y = property_value, 
        geom = "line", group = 1, 
        color = I("green"),  # color of the line
        size = I(2),         # width of line
        main = paste("Property Value Trend in", county_name),
        xlab = "Year Range", ylab = "Property Value") +
    scale_y_continuous(labels = scales::dollar_format(prefix = "$")) 
    
  
}

# Replace "Your County Name" with the actual county name you want to plot
selected_county <- "Dallas County"  #Can give any county name

# Plot Property Value Trend for the Selected County
property_value_plot <- plot_property_value(housing_df, selected_county)
print(property_value_plot)


```


```{r}
##Change in Mean Income Over Time
###slide 8

income_df_mean <- rename(income_df, Measure_Values = `Measure Values`)

income_df_mean <- mutate(income_df_mean, Measure_Values = as.numeric(Measure_Values))

meanincome20072011 <- income_df_mean %>%
  filter(Year == '2007-2011')

mean20072011 <- mean(meanincome20072011$Measure_Values)


meanincome20122016 <- income_df_mean %>%
  filter(Year == '2012-2016')

mean20122016 <- mean(meanincome20122016$Measure_Values)

meanincome20172021 <- income_df_mean %>%
  filter(Year == '2017-2021')

mean20172021 <- mean(meanincome20172021$Measure_Values)


##Change in Mean Income By Race 

###slide 9

income_df_mean_change_race <- income_df_mean %>% 
    group_by(`Race Origin`, Year) %>%
    summarize(mean(Measure_Values)
) 

income_df_mean_change_race
```


# Further Breakdown of Income on Home Value and Over Time

```{r}
# Using Overall Income by county alongside Most Recent 5 year sample


MHI_Cleaned <- income_df %>% 
    filter(`Race Origin` == "Total", `Measure Names` == "Estimate")

recent <- MHI_Cleaned %>% 
    filter(Year == "2017-2021", `Income Level` == "Median Household Income")

recent$MedianCountyValue <-as.numeric(housing_df$`Estimate2017-2021`)



#Impact of Income on Household value
plot1 <- recent %>%
    ggplot(aes(x = `Measure Values`, y = MedianCountyValue )) +
    geom_point(alpha = .3) +
    #add names using ggrepel limiting overlap
    geom_text_repel(aes(label=recent$`Geographic name`)) +
    #titles and caption
    labs(x = "Median Household Income",
         y = "Median Household Value",
         title = "Impact of Household Income on Home Value",
         caption = "Source: Iowa State Data Center") +
    theme_bw() +
    theme(
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold")
    ) +
    #make ticks look nice
    scale_y_continuous(labels = dollar_format(prefix = "$"),breaks = scales::pretty_breaks(n = 10)) +
    scale_x_continuous(labels = dollar_format(prefix = "$"), breaks = scales::pretty_breaks(n = 10))

plot1

#choropleth

data(county.regions)

cr <- county.regions

cr <- cr %>%
    filter(state.abb == "IA") %>%
    select(region, county.name) %>%
    arrange(county.name)

housing_dfPlot <- housing_df %>% mutate(region = cr$region) %>% 
    select(`Estimate2017-2021`,region)



# How have property values changed over time

MHI_Cleaned$Year <- as.factor(MHI_Cleaned$Year)

plot3 <- MHI_Cleaned %>% 
    filter(`Income Level` == "Median Household Income") %>% 
    ggplot(aes(x = Year, y = `Measure Values`)) + 
    geom_boxplot(aes(fill = Year)) + 
    theme_minimal() + 
    ggtitle("Recent Growth of Median Household Income") +
    xlab("Time Period") +
    ylab("Median Household Income") + 
    scale_y_continuous(labels = dollar_format(prefix = "$"),breaks = scales::pretty_breaks(n = 5)) + 
    labs(caption = "Source: Iowa State Data Center")

plot3

#What counties have the highest value

plot4_df <- county.regions %>% 
    filter(state.name == "iowa") 

plot4_df$value <-  housing_df$`Estimate2017-2021`



plot4 <- county_choropleth(plot4_df, state_zoom = "iowa") +
    scale_fill_brewer(palette = "YlOrRd")

plot4





```


#Q3) What Counties are similar based on traits such as Home Value, unemployment rate, poverty, racial makeup ect?
```{r}

# Loading in Other County Level Data and Transforing to get the desired measure. Using most recent timeframe

Unemployment<-read_xlsx("Data/Unemployment.xlsx")

DemoData<- read_xlsx("Data/Demographic Population.xlsx")

Poverty<- read_xlsx("Data/Poverty.xlsx")

Unemployment <- Unemployment %>%
  drop_na() %>%
  select(-Year, -`Margin of Error`)%>%
  rename(`Geographic name`=`Geographic Name`)

Poverty <- Poverty %>%
  rename(
    PovertyRate = `2017-2021`,
    `Geographic name` = `Goegraphic Name`) %>%
    mutate(`Geographic name` = paste0(`Geographic name`, " County"))


DemoData <- DemoData %>%
  rename( `Geographic name` = `Geographic Name`)

Dropped_housing_df <- housing_df %>%
  select(`Geographic name`, `Estimate2017-2021`) %>%
  rename(MedianHousingPrice = `Estimate2017-2021`)

Dropped_income_df <- income_df %>%
  filter(`Race Origin` == "Total",
         `Income Level` == "Median Household Income",
         Year == "2017-2021",
         `Measure Names` == "Estimate") %>%
  select(`Geographic name`, `Measure Values`) %>%
  rename(Income = `Measure Values`)

# Joining to one dataframe by county

CountyDF <- Unemployment %>%
  inner_join(DemoData, by = "Geographic name") %>%
  inner_join(Poverty, by = "Geographic name") %>%
  inner_join(Dropped_housing_df, by = "Geographic name")%>%
  inner_join(Dropped_income_df, by = "Geographic name")


data_for_clustering <- CountyDF %>% select(-`Geographic name`) 
# Scaling as K-means highly infuenced by outliers
scaled_data <- scale(data_for_clustering)

set.seed(123) 
wss <- map_dbl(1:10, function(k) {
  model <- kmeans(scaled_data, centers = k)
  model$tot.withinss
})

#Plot to look at where there could be a good number of clusters with valuable info

elbow_plot <- data.frame(
  k = 1:10,
  wss = wss
)

ggplot(elbow_plot, aes(x = k, y = wss)) +
  geom_line() +
  geom_point() +
  ggtitle("Elbow Method") +
  ylab("Within-cluster Sum of Squares") +
  xlab("Number of Clusters")



```


```{r}
# Determined 5 based off above graph

k_optimal <- 5

set.seed(123)  
kmeans_result <- kmeans(scaled_data, centers = k_optimal)


CountyDF$cluster <- kmeans_result$cluster

## Looking at summaries of each cluster

summary_table <- CountyDF %>%
  select(-`Geographic name`) %>%
  group_by(cluster) %>%
  summarise_all(mean, na.rm = TRUE) %>%
  mutate(
    `Unemployment rate` = percent(`Unemployment rate` / 100, accuracy = 0.01),
    Population = as.integer(Population),
    across(4:10, ~percent(.x, accuracy = 0.01)),
    PovertyRate = percent(PovertyRate / 100, accuracy = 0.01),
    MedianHousingPrice = dollar(MedianHousingPrice, accuracy = 1),
    Income = dollar(Income, accuracy = 1)
  ) %>%
  left_join(CountyDF %>% group_by(cluster) %>% summarise(NumObservations = n()), by = "cluster")

print(summary_table)

```

```{r}
#Loading Iowa Data and joining to built dataframe

iowa_shape <- counties(state = "IA")

iowacountydata <- left_join(iowa_shape, CountyDF, by = c("NAMELSAD" = "Geographic name"))

#Defining Color Palette for our clusters that went from 6 to 5

color_palette <- c("blue", "red", "green", "yellow", "purple", "orange")

ggplot(data = iowacountydata) + 
  geom_sf(aes(fill = as.factor(cluster)), color = "white") +
  scale_fill_manual(values = color_palette) +
  theme_minimal() +
  labs(fill = "cluster", title = "Cluster Assignments by County") +
  theme(legend.position = "right")

cluster_plot <- ggplot(data = iowacountydata) + 
  geom_sf(aes(fill = as.factor(cluster)), color = "white") +
  scale_fill_manual(values = c("blue", "red", "green", "yellow", "purple", "orange")) +
  theme_minimal() +
  labs(fill = "Cluster", title = "Cluster Assignments by County") +
  theme(legend.position = "right")


```


#Q4)How well does median income predict median housing value and vice versa?

```{r}
# Merge based on 'Geographic Name' column
merged_data <- merge(housing_df, income_df, by.x = "Geographic name", by.y = "Geographic name")

# Filter out rows where Income Level is Per Capita Income
filtered_data_sc <- merged_data[merged_data$`Income Level` != "Per Capita Income", ]

# Visualization: Scatter plot with facets based on Income Level
plot <- qplot(data = filtered_data_sc, x = `Measure Values`, y = `Estimate2017-2021`, geom = "point", color = `Income Level`, facets = . ~ `Income Level`) +
  labs(title = "Median Income vs. Median Housing Value", x = "Median Income", y = "Housing Value") +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
  scale_x_continuous(labels = scales::dollar_format(prefix = "$")) +
  theme(axis.text = element_text(size = 6)) +
  theme(axis.text = element_text(color = "black"))

plot

```

#Q5)Does the racial makeup of a county affect the median housing value in that    area?
```{r}
# Filter out rows where Race Origin is "Total"
filtered_data <- merged_data[merged_data$`Race Origin` != "Total", ]

# Create a horizontal violin plot using qplot

plot2 <- qplot(data = filtered_data, x = `Estimate2017-2021`, y = `Race Origin`, fill = `Race Origin`,
              geom = "violin", main = "Effect of Racial Makeup on Median Housing Value for 2017 - 2021",
              ylab = "Race", xlab = "Median Housing Value") +
        scale_fill_discrete(name = "Race") +
        scale_x_continuous(labels = scales::dollar_format(prefix = "$")) +  # Format X-axis as currency
        theme_minimal() +
        theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 8),
              axis.text.x = element_text(angle = 0, hjust = 1, size = 8),  # Adjust y-axis 
              axis.title.x = element_text(color = "black"),  # Set x-axis title color to black
              axis.title.y = element_text(color = "black"),  # Set y-axis title color to black
              axis.text = element_text(color = "black"),     # Set axis text color to black
              legend.text = element_text(color = "black"),
              legend.title = element_text(color = "black"),
              panel.background = element_rect(fill = "lightblue"),# Set background color
              plot.title = element_text(color = "black", size = 14, face = "bold"))  # Set title color and size) +
 
 
plot2


```


